---
title: "TO404 Homework Assignment 1"
author: "Richard Einhorn"
date: "Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r, message=F, error=F}
#install.packages("nycflights13")
library(nycflights13)  # import data
library(ggplot2)  # used in visualizations
library(dplyr)  # used in data frame manipulations
nyc <- nycflights13::flights  # assign data to nyc data frame
```

Let's take a look at the data:
```{r}
head(nyc)
```


# Basic Exploration

Let’s first do some simple exploration of this data.

## How many airlines are there?
```{r}
length(table(na.omit(nyc$carrier)))
```
There are 16 airlines in total.

## How many flights there were by the airline with code OO?
```{r}
nrow(na.omit(nyc[nyc$carrier == "OO",]))
```
There were a total of 29 flights by the airline with code OO. 

## How long is the shortest flight out of any NYC airport??
```{r}
min(nyc$air_time, na.rm = TRUE)
```
The shortest flight recorded took 20 minutes.

## How many flights were there by United Airlines on Jan 12th 2013?
```{r}
nrow(na.omit(nyc[nyc$carrier == "UA" & nyc$month == '1' & nyc$day == '12',]))
```
There were 110 flights by UA on Jan 12th 2013.

# Arrival Delay

Lets focus on Arrival Delay.

## What was the average arrival delay?
```{r}
mean(na.omit(nyc[nyc$month == '1',]$arr_delay))
```
The average arrival delay for all airports and all airlines combined in Jan 2013 was equal to about 6 minutes.

## What was the median arrival delay?
```{r}
median(na.omit(nyc[nyc$month == '1',]$arr_delay))
```
The median arrival delay for all airports and all airlines combined in Jan 2013 was equal to about -3 minutes, meaning that exactly 50% of flights arrived at least 3 minutes early.

> We see that the average flight delay was around 6 minutes indicating that, overall, flights from NYC airports were efficient (assuming that passengers do not seriously mind a delay of a few minutes). The difference between the mean and the median suggests that the data is skewed (to the right), meaning that there is a few flights which are massively delayed hence have a large impact on the mean. It could be argued that passengers are probably more interested in the median flight arrival delay as most flights (i.e. more than 50%) do not have any arrival delay (in fact, the majority of flights in the dataset arrive early).

# Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned.

## What is the average arrival delay by airline?
```{r}
nyc$carrier <- as.factor(nyc$carrier) # ensuring that airline column in data set
# is coded as a factor

carrier_delay = tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)
knitr::kable(carrier_delay, col.names="Average arrival delay (in minutes)", digits = 2)
```

```{r, echo=F}
colour <- ifelse(carrier_delay < 0, 'green', 'red')
barplot(carrier_delay, xlab = 'Airline Code', ylab = 'Average arrival delay (in minutes)', main = 'Comparison of Average Arrival Delay by Airline', col = colour)

```

We can see that only two airlines (AS and HA) do not, on average, arrive with a delay. If we are looking for the airline with the least divergence from the expected arrival time, we find that AA performs best with a minimum absolute average arrival delay of .36 minutes.

## Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there any airlines that have a negative average arrival delay?

We find that `{r} which.max(carrier_delay)` has the highest average arrival delay (equal to approx. 22 minutes). If we consider airlines that actually have a negative average arrival delay (i.e. whose flights usually arrive before the scheduled time), we find that `{r} which.min(carrier_delay)` has the smallest average arrival delay (equal to approx. -10 minutes). From those airlines that, on average, arrive delayed, `{r} which.min(carrier_delay[carrier_delay>0])` has the smallest average arrival delay.

Finally, as discussed above, two airlines have negative average arrival delays: `{r} which(carrier_delay < 0)`.

# Air Gain

Let's find out whether airlines actually make up time when on air.

## Explore airgain data

### Summary statistics
```{r}
nyc$air_gain = nyc$dep_delay - nyc$arr_delay
summary(nyc$air_gain)

hist(nyc$air_gain, 100, probability = TRUE, main = "Distribution of air gain", xlab = "Air Gain")
```

### Air Gain by Airline
```{r, warning=F}
ggplot(data = nyc, aes(x = carrier, y = air_gain)) +
  geom_jitter(alpha = 0.4) +
  labs(
    title = "Air Gain distribution by Airline",
    x = "Airline",
    y = "Air Gain"
  )
```

## Questions regarding air gain data

### Do airlines actually gain any time when in air on average?
```{r}
mean(na.omit(nyc$air_gain))
```
Airlines, on average, do gain time while in the air (approx. 6 minutes in the mean).

### Which airlines do a better job, which do a worse job?
```{r}
sort(tapply(nyc$air_gain, nyc$carrier, mean, na.rm = TRUE))
```
The above table shows how airlines perform on air gain (sorted from worst to best). Note that a negative value means that the airline's flights, on average, increase the delay while in the air (airlines F9 and FL). This is visualized in the below bar plot:

```{r}
barplot(sort(tapply(nyc$air_gain, nyc$carrier, mean, na.rm = TRUE)), main="Average air gain by airline", xlab="Airline Code", ylab = "Average Air Gain (in minutes)")
```
